---
title: "Module 4 Multivariate Statistics"
author: "Eric R. Schuler, Ph.D."
date: "`r Sys.Date()`"
format: gfm
---

In this module, we will cover how to run a:

- MANOVA
- Linear Discriminant Analysis
- Factor Analysis
- Canonical Correlation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

list.of.packages <- c("tidyverse","car","yhat","lmtest","forcats","psych","apaTables","sjstats","janitor","dplyr","ggplot2","gridExtra","sandwich","DescTools","performance","see")
#If you do not already have the package installed, the package is retained
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
#Install packages that are not previously installed
if(length(new.packages)) install.packages(new.packages,repos = "http://cran.us.r-project.org")

options(scipen = 999)

#set working directory
setwd("C:/Users/eschuler/Desktop/r short course")

#Load packages
library(tidyverse)
library(broom)
```

#Import data
```{r}
hs1939 <- read.csv("hs1939_cleaned.csv")
hs1939 <- hs1939[-5,] #dropping case 5 that has some missing
```

we will change sex to a factor
```{r}
hs1939$sex <- factor(hs1939$sex)
janitor::tabyl(hs1939$sex)
```

We will now make the variable grade into an ordinal variable
```{r}
hs1939$grade <- ordered(hs1939$grade, labels = c("7th grade", "8th grade"))
```

Make school a factor
```{r}
hs1939$school <- factor(hs1939$school)
```

# MANOVA
```{r}
manova_res <- manova(cbind(cubes,visual_perception) ~ school, data = hs1939)
summary(manova_res)
```

Follow-up ANOVAs
```{r}
summary.aov(manova_res)
```

# Linear Discriminant Analysis (lda)
```{r}
library(MASS)
```

Run an lda
```{r}
lda_res <- lda(school ~ cubes + lozenges + visual_perception, data = hs1939)
```

View the results
```{r}
lda_res
```

We can then plot the model
```{r}
plot(lda_res)
```

Factor Analysis
Code adapted from: https://rpubs.com/pjmurphy/758265

We will be using the psych package for this.
```{r}
library(psych)
```

Holzinger and Swineford 1939 data but dropping demographics and variables we created
```{r}
hs<-hs1939[,-c(1:6,16,17)]
```

Running a KMO test
```{r}
KMO(hs)
```

Bartlett's test of sphericity
```{r}
cortest.bartlett(hs)
```

Dimensionality

We will use a couple methods

Eigenvalue greater than 1
```{r}
ev <- eigen(cor(hs))
ev$values
```
Here we can see three factors

Let's examine a scree plot
  If pc=TRUE it will be for principle components analysis (PCA)
```{r}
scree(hs, pc=FALSE)
```

Better option is for parallel analysis
  If running a PCA, use "pc" rather than "fa"
```{r}
fa.parallel(hs,fa="fa")
```

Another good option is vss which also runs minimum average partial
```{r}
vss(hs)
```

factor analysis

we will set the number of factors
```{r}
Nfac <- 3
```

First we need to activate another package
```{r}
library(GPArotation)
```
  
Here we have the dataset, number of factors, and we will use promax, an oblique rotation, if we found a low correlation, we would change promax to varimax to have it be an orthogonal rotation which makes it easier to interpret.
```{r}
fa.obl <- fa(cov(hs),fm="pa",Nfac,rotate="promax")
```

```{r}
fa.obl
```
Print the coefficients (path coefficients, pattern coefficients, uniqueness, and communality)
```{r}
print(fa.obl, digits=2, cutoff=.3, sort=TRUE)
```
plot factor 1 by 2
```{r}
load1 <- fa.obl$loadings[,1:2]
plot(load1,type="n")
text(load1,labels=names(hs),cex=.7)
```

plot factor 1 by 3
```{r}
load2 <- fa.obl$loadings[,c(1,3)]
plot(load2,type="n")
text(load2,labels=names(hs),cex=.7)
```

plot factor 2 by 3
```{r}
load3 <- fa.obl$loadings[,c(2,3)]
plot(load3,type="n")
text(load3,labels=names(hs),cex=.7)
```

We can also inspect a diagram
```{r}
loads <- fa.obl$loadings
fa.diagram(loads)
```

We can then run internal reliabilty

First we subset by factor
```{r}
f1 <- hs[,c("visual_perception","cubes","lozenges")]
f2 <- hs[,c("paragraph_completion","sentence_completion","word_meaning")]
f3 <- hs[,c("speeded_addition","speeded_counting","speeded_discrimination")]
```


Then reliability

Factor 1
```{r}
psych::alpha(f1)
```

Factor 2
```{r}
psych::alpha(f2)
```

Factor 3
```{r}
psych::alpha(f3)
```

# Cannonical Correlation

Code adapted from: https://stats.oarc.ucla.edu/r/dae/canonical-correlation-analysis/

```{r}
library(ggplot2)
library(GGally)
library(CCA)
```



We will now make a group of predictors
```{r}
pred <- hs1939[,10:12]
```

And a group of outcomes
```{r}
out <- hs1939[,7:9]
```

View the pairs of predictor variables
```{r}
ggpairs(pred)
```

View the pairs of outcome variables
```{r}
ggpairs(out)
```
Print out correlations of the sets
```{r}
matcor(pred,out)
```

run a canonical correlation
```{r}
cc_res <- cc(pred,out)
```

Look at the canonical correlation
```{r}
cc_res$cor
```

raw canonical correlation
```{r}
cc_res[3:4]
```


compute and displate the canonical loadings
```{r}
cc2 <- comput(pred, out, cc_res)

# display canonical loadings
cc2[3:6]
```

We can test the number of canonical dimensions
```{r}
rho <- cc_res$cor
```


We will then define the number of observations, and variables in each set
```{r}
n <- dim(pred[1])
p <- length(pred)
o <- length(out)
```

Calculate the p-values
```{r}
library(CCP)
p.asym(rho,n,p,o, tstat="Wilks")
p.asym(rho,n,p,o, tstat="Hotelling")
p.asym(rho,n,p,o, tstat="Pillai")
```

Now we can standardize the coefficients for interpretation

Pred
```{r}
s1 <- diag(sqrt(diag(cov(pred))))
s1 %*% cc_res$xcoef
```

Out
```{r}
s1 <- diag(sqrt(diag(cov(out))))
s1 %*% cc_res$xcoef
```


Resources:

- MANOVA: http://www.sthda.com/english/wiki/manova-test-in-r-multivariate-analysis-of-variance
- Factor Analysis: https://rpubs.com/pjmurphy/758265
- DDA: http://www.sthda.com/english/articles/36-classification-methods-essentials/146-discriminant-analysis-essentials-in-r/#:~:text=Discriminant%20analysis%20is%20used%20to,and%2For%20categorical%20predictor%20variables. 
- CCA: https://stats.idre.ucla.edu/r/dae/canonical-correlation-analysis/